---
title: "LearnR3"
author: "Ingrid M. Wagnon"
format:
  html:
    number-sections: true
    markdown:
      hardline-breaks: true
---

Note about the YAML : Very sensitive to spacing and tabulation.

```{r setup}
library(tidyverse)
library(here)
library(snakecase)
```


# Importing Raw Data

## Import CGM

Just about GitHub : If you want it private you can add a private argument.

When you load data from a CSV file into R with read_csv(), the function will try to guess what the data types are for each column. Usually it guesses accurately.
The argument show_col_types = FALSE tells R to just guess and to not give a warning message.
Otherwise we could explicitly tell read_csv() what the column names and data types are using the argument col_types. This is a very tedious, but often very powerful and useful step depending on the issues your data may have.

The column names are not in a format that we can easily use. For example, the first column is called "Device Timestamp" and the space in the name will make it harder to work with. So, we will need to fix the column names up to follow a common style and that is easier to code with. That style is snake_case that uses _ for spaces and is all lower case.
The function from snakecase that we want to use in the name_repair argument is the to_snake_case() function. To convert the variable names to snake case, we would add name_repair = to_snake_case to the code

The n_max argument to only load in the first 100 rows

If necessary (not shown in this example) you can use col_select = arguments in read_csv() to import only the data you need.

```{r importing csv}
cgm_101 <- here("data-raw/dime/cgm/101.csv") |>
  read_csv(
    show_col_types = FALSE,
    name_repair = to_snake_case,
    n_max = 100
  )
```


## Exercise: Import 101's sleep data

```{r importing csv exercise}
sleep_101 <- here("data-raw/dime/sleep/101.csv") |>
  read_csv(
    show_col_types = FALSE,
    name_repair = to_snake_case,
    n_max = 101
  )
sleep_101
```


# Creating Functions

Basic structure of a function:

name <- function(argument1, argument2) {
    # body of function
    output <- ... code ....
    return(output)
}

## Making a function to add numbers

```{r example of adding numbers}
#' Add two numbers together.
#'
#' @param num1 A number here.
#' @param num2 A number here.
#'
#' @return Returns the sum of the two numbers.
#'

add_numbers <- function(num1, num2) {
  added <- num1 + num2
  return(added)
}

add_numbers(1, 2)
```

## Import 101's cgm data with a function

From how we import our data we create a function

It's good practice to not hard code the use of here() within a function. Instead, it’s good design to give functions a full file path that it can use internally. Then when we use the function, we would use here() with the correct path in the function argument. 

Then we simplify things internally by renaming cgm_101 to simply cgm, since we would like to be able to import other participant CGM data later.

```{r create cgm function}
#' Import one participants CGM data from the DIME dataset.
#'
#' @param file_path Path to the CGM dataset.
#'
#' @returns Outputs a data frame/tibble.
#'
import_cgm <- function(file_path) {
  cgm <- file_path |>
    read_csv(
      show_col_types = FALSE,
      name_repair = to_snake_case,
      n_max = 100
    )
  return(cgm)
}
```

```{r test 1 cgm}
here("data-raw/dime/cgm/101.csv") |>
  import_cgm()
```

```{r test 2 cgm}
here("data-raw/dime/cgm/102.csv") |>
  import_cgm()
```

# Exercise : Create sleep function

name <- function(argument1, argument2) {
    # body of function
    output <- ... code ....
    return(output)
}

```{r create sleep function}
#' Import one participants SLEEP data from the DIME dataset.
#'
#' @param file_path Path to the CGM dataset.
#'
#' @returns Outputs a data frame/tibble.
#'
import_sleep <- function(file_path) {
  sleep <- file_path |>
    read_csv(
      show_col_types = FALSE,
      name_repair = to_snake_case,
      n_max = 101
    )
  return(sleep)
}
```

```{r test 1 sleep}
here("data-raw/dime/sleep/101.csv") |>
  import_sleep()
```

```{r test 2 sleep}
here("data-raw/dime/sleep/102.csv") |>
  import_sleep()
```

# Import DIME data function : Make function more general

Specifically call the package from which the function you are using comes from 

To make sure your code runs smoothly use the function usethis::use_package("library called") in the console which should add the necessary packages to the DESCRIPTION file.
For tidyverse use : usethis::use_package("tidyverse", type = "Depends")

```{r import_dime function}
#' Import data from the DIME study dataset.
#'
#' @param file_path Path to the CSV file.
#' @param n Optional parameter that limits the number of rows. Default is 100.
#'
#' @returns A data frame.
#'
import_dime <- function(file_path, n = 100) {
  data <- file_path |>
    readr::read_csv(
      show_col_types = FALSE,
      name_repair = snakecase::to_snake_case,
      n_max = n
    )
  return(data)
}
```

```{r}
here("data-raw/dime/sleep/101.csv") |>
  import_dime()
```
